\section{Bibliography}\label{bibliography}

This page provides a comprehensive collection of references, external
resources, and further reading for the Chimera DSP Wiki.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Table of Contents}{Table of Contents}}\label{table-of-contents}

\begin{itemize}
\tightlist
\item
  \hyperref[textbooks]{Textbooks}
\item
  \hyperref[research-papers]{Research Papers}
\item
  \hyperref[standards-and-specifications]{Standards and Specifications}
\item
  \hyperref[online-resources]{Online Resources}
\item
  \hyperref[signal-databases-and-references]{Signal Databases and
  References}
\item
  \hyperref[software-tools]{Software Tools}
\item
  \hyperref[organizations-and-institutions]{Organizations and
  Institutions}
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Textbooks}{Textbooks}}\label{textbooks}

\subsubsection{Digital Communications \& Signal
Processing}\label{digital-communications-signal-processing}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Proakis, J. G., \& Salehi, M.} (2007). \emph{Digital
  Communications} (5th ed.). McGraw-Hill.

  \begin{itemize}
  \tightlist
  \item
    Comprehensive coverage of modulation, coding, and detection theory
  \item
    Recommended for: Parts IV-V
  \end{itemize}
\item
  \textbf{Haykin, S.} (2013). \emph{Communication Systems} (5th ed.).
  Wiley.

  \begin{itemize}
  \tightlist
  \item
    Systems-level approach to analog and digital communications
  \item
    Recommended for: Parts II-IV
  \end{itemize}
\item
  \textbf{Sklar, B.} (2001). \emph{Digital Communications: Fundamentals
  and Applications} (2nd ed.). Prentice Hall.

  \begin{itemize}
  \tightlist
  \item
    Practical emphasis with detailed examples
  \item
    Recommended for: Parts IV-VI
  \end{itemize}
\item
  \textbf{Tse, D., \& Viswanath, P.} (2005). \emph{Fundamentals of
  Wireless Communication}. Cambridge University Press.

  \begin{itemize}
  \tightlist
  \item
    Modern wireless systems, MIMO, multicarrier
  \item
    Recommended for: Parts VI-VII
  \end{itemize}
\item
  \textbf{Goldsmith, A.} (2005). \emph{Wireless Communications}.
  Cambridge University Press.

  \begin{itemize}
  \tightlist
  \item
    Advanced topics in channel modeling and capacity
  \item
    Recommended for: Parts III, VI-VII
  \end{itemize}
\item
  \textbf{Oppenheim, A. V., \& Schafer, R. W.} (2009).
  \emph{Discrete-Time Signal Processing} (3rd ed.). Prentice Hall.

  \begin{itemize}
  \tightlist
  \item
    Essential DSP fundamentals
  \item
    Recommended for: Background reading
  \end{itemize}
\end{enumerate}

\subsubsection{Channel Coding Theory}\label{channel-coding-theory}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  \textbf{Richardson, T., \& Urbanke, R.} (2008). \emph{Modern Coding
  Theory}. Cambridge University Press.

  \begin{itemize}
  \tightlist
  \item
    In-depth treatment of LDPC and turbo codes
  \item
    Recommended for: Part V
  \end{itemize}
\item
  \textbf{Moon, T. K.} (2005). \emph{Error Correction Coding:
  Mathematical Methods and Algorithms}. Wiley.

  \begin{itemize}
  \tightlist
  \item
    Comprehensive coding theory reference
  \item
    Recommended for: Part V
  \end{itemize}
\item
  \textbf{Lin, S., \& Costello, D. J.} (2004). \emph{Error Control
  Coding} (2nd ed.). Prentice Hall.

  \begin{itemize}
  \tightlist
  \item
    Classic coding theory textbook
  \item
    Recommended for: Part V
  \end{itemize}
\end{enumerate}

\subsubsection{Electromagnetic Theory \&
Propagation}\label{electromagnetic-theory-propagation}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{9}
\tightlist
\item
  \textbf{Griffiths, D. J.} (2017). \emph{Introduction to
  Electrodynamics} (4th ed.). Cambridge University Press.

  \begin{itemize}
  \tightlist
  \item
    Maxwell\textquotesingle s equations and wave propagation
  \item
    Recommended for: Part I
  \end{itemize}
\item
  \textbf{Balanis, C. A.} (2016). \emph{Antenna Theory: Analysis and
  Design} (4th ed.). Wiley.

  \begin{itemize}
  \tightlist
  \item
    Comprehensive antenna engineering reference
  \item
    Recommended for: Part I
  \end{itemize}
\item
  \textbf{Stutzman, W. L., \& Thiele, G. A.} (2012). \emph{Antenna
  Theory and Design} (3rd ed.). Wiley.

  \begin{itemize}
  \tightlist
  \item
    Practical antenna design
  \item
    Recommended for: Part I
  \end{itemize}
\item
  \textbf{Parsons, J. D.} (2000). \emph{The Mobile Radio Propagation
  Channel} (2nd ed.). Wiley.

  \begin{itemize}
  \tightlist
  \item
    Multipath, fading, and channel modeling
  \item
    Recommended for: Parts II-III
  \end{itemize}
\end{enumerate}

\subsubsection{Quantum Biology (Part
VIII)}\label{quantum-biology-part-viii}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{13}
\tightlist
\item
  \textbf{Penrose, R.} (1989). \emph{The Emperor\textquotesingle s New
  Mind}. Oxford University Press.

  \begin{itemize}
  \tightlist
  \item
    Origins of Orch-OR theory
  \item
    Recommended for: Part VIII
  \end{itemize}
\item
  \textbf{Al-Khalili, J., \& McFadden, J.} (2014). \emph{Life on the
  Edge: The Coming Age of Quantum Biology}. Broadway Books.

  \begin{itemize}
  \tightlist
  \item
    Popular science introduction to quantum biology
  \item
    Recommended for: Part VIII
  \end{itemize}
\item
  \textbf{Hameroff, S., \& Penrose, R.} (2014). ``Consciousness in the
  universe: A review of the `Orch OR' theory.'' \emph{Physics of Life
  Reviews}, 11(1), 39-78.

  \begin{itemize}
  \tightlist
  \item
    Peer-reviewed Orch-OR overview
  \item
    Recommended for: Part VIII
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Research Papers}{Research Papers}}\label{research-papers}

\subsubsection{Foundational Papers}\label{foundational-papers}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{16}
\tightlist
\item
  \textbf{Shannon, C. E.} (1948). ``A Mathematical Theory of
  Communication.'' \emph{Bell System Technical Journal}, 27(3), 379-423.

  \begin{itemize}
  \tightlist
  \item
    Foundation of information theory
  \item
    DOI: 10.1002/j.1538-7305.1948.tb01338.x
  \end{itemize}
\item
  \textbf{Viterbi, A. J.} (1967). ``Error bounds for convolutional codes
  and an asymptotically optimum decoding algorithm.'' \emph{IEEE
  Transactions on Information Theory}, 13(2), 260-269.

  \begin{itemize}
  \tightlist
  \item
    Viterbi algorithm for decoding convolutional codes
  \item
    DOI: 10.1109/TIT.1967.1054010
  \end{itemize}
\item
  \textbf{Berrou, C., Glavieux, A., \& Thitimajshima, P.} (1993). ``Near
  Shannon limit error-correcting coding and decoding: Turbo-codes.''
  \emph{Proceedings of ICC\textquotesingle93}, 1064-1070.

  \begin{itemize}
  \tightlist
  \item
    Introduction of turbo codes
  \item
    DOI: 10.1109/ICC.1993.397441
  \end{itemize}
\item
  \textbf{Gallager, R. G.} (1962). ``Low-density parity-check codes.''
  \emph{IRE Transactions on Information Theory}, 8(1), 21-28.

  \begin{itemize}
  \tightlist
  \item
    Original LDPC codes paper
  \item
    DOI: 10.1109/TIT.1962.1057683
  \end{itemize}
\end{enumerate}

\subsubsection{Modern Wireless Systems}\label{modern-wireless-systems}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{20}
\tightlist
\item
  \textbf{Foschini, G. J., \& Gans, M. J.} (1998). ``On limits of
  wireless communications in a fading environment when using multiple
  antennas.'' \emph{Wireless Personal Communications}, 6(3), 311-335.

  \begin{itemize}
  \tightlist
  \item
    MIMO capacity analysis
  \item
    DOI: 10.1023/A:1008889222784
  \end{itemize}
\item
  \textbf{Telatar, E.} (1999). ``Capacity of multi-antenna Gaussian
  channels.'' \emph{European Transactions on Telecommunications}, 10(6),
  585-595.

  \begin{itemize}
  \tightlist
  \item
    Theoretical MIMO capacity
  \item
    DOI: 10.1002/ett.4460100604
  \end{itemize}
\end{enumerate}

\subsubsection{THz and Quantum Biology}\label{thz-and-quantum-biology}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{22}
\tightlist
\item
  \textbf{Fr\"{o}hlich, H.} (1968). ``Long-range
  coherence and energy storage in biological systems.''
  \emph{International Journal of Quantum Chemistry}, 2(5), 641-649.

  \begin{itemize}
  \tightlist
  \item
    Biological quantum coherence theory
  \item
    DOI: 10.1002/qua.560020505
  \end{itemize}
\item
  \textbf{Craddock, T. J., et al.} (2014). ``Anesthetic alterations of
  collective terahertz oscillations in tubulin correlate with clinical
  potency.'' \emph{Scientific Reports}, 4, 4977.

  \begin{itemize}
  \tightlist
  \item
    Microtubule THz resonances and anesthesia
  \item
    DOI: 10.1038/srep04977
  \end{itemize}
\item
  \textbf{Frey, A. H.} (1962). ``Human auditory system response to
  modulated electromagnetic energy.'' \emph{Journal of Applied
  Physiology}, 17(4), 689-692.

  \begin{itemize}
  \tightlist
  \item
    Original Frey effect discovery
  \item
    DOI: 10.1152/jappl.1962.17.4.689
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Standards and Specifications}{Standards and Specifications}}\label{standards-and-specifications}

\subsubsection{ITU-R (International Telecommunication Union -
Radiocommunication
Sector)}\label{itu-r-international-telecommunication-union---radiocommunication-sector}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{25}
\tightlist
\item
  \textbf{ITU-R P.618-13} (2017). ``Propagation data and prediction
  methods required for the design of Earth-space telecommunication
  systems.''

  \begin{itemize}
  \tightlist
  \item
    Rain attenuation models, atmospheric absorption
  \item
    URL: \url{https://www.itu.int/rec/R-REC-P.618/en}
  \end{itemize}
\item
  \textbf{ITU-R P.676-12} (2019). ``Attenuation by atmospheric gases and
  related effects.''

  \begin{itemize}
  \tightlist
  \item
    Oxygen and water vapor absorption
  \item
    URL: \url{https://www.itu.int/rec/R-REC-P.676/en}
  \end{itemize}
\item
  \textbf{ITU-R P.530-17} (2017). ``Propagation data and prediction
  methods required for the design of terrestrial line-of-sight
  systems.''

  \begin{itemize}
  \tightlist
  \item
    Multipath and rain fading for terrestrial links
  \item
    URL: \url{https://www.itu.int/rec/R-REC-P.530/en}
  \end{itemize}
\end{enumerate}

\subsubsection{3GPP (3rd Generation Partnership
Project)}\label{gpp-3rd-generation-partnership-project}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{28}
\tightlist
\item
  \textbf{3GPP TS 38.211} (2021). ``NR; Physical channels and
  modulation.''

  \begin{itemize}
  \tightlist
  \item
    5G NR physical layer specification
  \item
    URL: \url{https://www.3gpp.org/ftp/Specs/archive/38_series/38.211/}
  \end{itemize}
\item
  \textbf{3GPP TS 36.212} (2020). ``LTE; Multiplexing and channel
  coding.''

  \begin{itemize}
  \tightlist
  \item
    LTE turbo coding and multiplexing
  \item
    URL: \url{https://www.3gpp.org/ftp/Specs/archive/36_series/36.212/}
  \end{itemize}
\end{enumerate}

\subsubsection{IEEE Standards}\label{ieee-standards}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{30}
\tightlist
\item
  \textbf{IEEE 802.11-2020}. ``Wireless LAN Medium Access Control (MAC)
  and Physical Layer (PHY) Specifications.''

  \begin{itemize}
  \tightlist
  \item
    WiFi standards (OFDM, MIMO, QAM)
  \item
    URL: \url{https://standards.ieee.org/standard/802_11-2020.html}
  \end{itemize}
\item
  \textbf{IEEE 802.16-2017}. ``Air Interface for Broadband Wireless
  Access Systems.''

  \begin{itemize}
  \tightlist
  \item
    WiMAX/Broadband wireless (OFDMA, AMC)
  \item
    URL: \url{https://standards.ieee.org/standard/802_16-2017.html}
  \end{itemize}
\end{enumerate}

\subsubsection{DVB (Digital Video
Broadcasting)}\label{dvb-digital-video-broadcasting}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{32}
\tightlist
\item
  \textbf{ETSI EN 302 307-1} (2014). ``DVB-S2: Framing structure,
  channel coding and modulation.''

  \begin{itemize}
  \tightlist
  \item
    Satellite broadcasting standard (LDPC, 8PSK, 16APSK, 32APSK)
  \item
    URL: \url{https://www.etsi.org/deliver/etsi_en/302300_302399/30230701/01.04.01_60/en_30230701v010401p.pdf}
  \end{itemize}
\end{enumerate}

\subsubsection{GPS/GNSS}\label{gpsgnss}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{33}
\tightlist
\item
  \textbf{IS-GPS-200M} (2021). ``Interface Specification: Navstar GPS
  Space Segment/Navigation User Segment Interfaces.''

  \begin{itemize}
  \tightlist
  \item
    GPS signal structure (L1 C/A, L2C, L5)
  \item
    URL: \url{https://www.gps.gov/technical/icwg/IS-GPS-200M.pdf}
  \end{itemize}
\item
  \textbf{Galileo OS SIS ICD} (2021). ``European GNSS (Galileo) Open
  Service Signal In Space Interface Control Document.''

  \begin{itemize}
  \tightlist
  \item
    Galileo signal specifications
  \item
    URL: \url{https://www.gsc-europa.eu/sites/default/files/sites/all/files/Galileo_OS_SIS_ICD_v2.0.pdf}
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Online Resources}{Online Resources}}\label{online-resources}

\subsubsection{Educational Websites}\label{educational-websites}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{35}
\tightlist
\item
  \textbf{RF Café} - \url{https://www.rfcafe.com/}

  \begin{itemize}
  \tightlist
  \item
    RF/microwave calculators, formulas, and references
  \item
    Topics: Link budgets, antenna gain, dBm conversions
  \end{itemize}
\item
  \textbf{Wireless Pi} - \url{https://wirelesspi.com/}

  \begin{itemize}
  \tightlist
  \item
    Educational tutorials on wireless communications
  \item
    Topics: Modulation, MIMO, OFDM
  \end{itemize}
\item
  \textbf{DSP Related} - \url{https://www.dsprelated.com/}

  \begin{itemize}
  \tightlist
  \item
    DSP tutorials, blogs, and forums
  \item
    Topics: Filter design, FFT, IQ modulation
  \end{itemize}
\item
  \textbf{GNURadio Tutorials} - \url{https://wiki.gnuradio.org/index.php/Tutorials}

  \begin{itemize}
  \tightlist
  \item
    Hands-on SDR tutorials
  \item
    Topics: Practical DSP implementation
  \end{itemize}
\item
  \textbf{Sharetechnote} - \url{https://www.sharetechnote.com/}

  \begin{itemize}
  \tightlist
  \item
    Comprehensive wireless system explanations
  \item
    Topics: LTE, 5G NR, WiFi physical layer
  \end{itemize}
\end{enumerate}

\subsubsection{Video Courses}\label{video-courses}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{40}
\tightlist
\item
  \textbf{MIT OpenCourseWare: Digital Communication Systems} - \url{https://ocw.mit.edu/courses/6-450-principles-of-digital-communications-i-fall-2006/}

  \begin{itemize}
  \tightlist
  \item
    Free MIT course on digital communications
  \item
    Lectures, notes, and problem sets
  \end{itemize}
\item
  \textbf{Stanford Engineering Everywhere: Digital Communications} - \url{https://see.stanford.edu/}

  \begin{itemize}
  \tightlist
  \item
    Stanford\textquotesingle s online courses
  \item
    Topics: Information theory, coding
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Signal Databases and References}{Signal Databases and References}}\label{signal-databases-and-references}

\subsubsection{Signal Identification}\label{signal-identification}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{42}
\tightlist
\item
  \textbf{Signal Identification Wiki (sigidwiki)} - \url{https://www.sigidwiki.com/wiki/Signal_Identification_Guide}

  \begin{itemize}
  \tightlist
  \item
    Comprehensive database of RF signals
  \item
    Useful for: Identifying real-world modulation schemes
  \item
    Topics: Amateur radio, military, satellite, broadcast signals
  \end{itemize}
\item
  \textbf{RadioReference.com} - \url{https://www.radioreference.com/}

  \begin{itemize}
  \tightlist
  \item
    Frequency allocations and trunked radio systems
  \item
    Topics: Public safety, aviation, maritime communications
  \end{itemize}
\end{enumerate}

\subsubsection{GNSS/GPS Resources}\label{gnssgps-resources}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{44}
\tightlist
\item
  \textbf{Navipedia (ESA)} - \url{https://gssc.esa.int/navipedia/}

  \begin{itemize}
  \tightlist
  \item
    European Space Agency\textquotesingle s GNSS encyclopedia
  \item
    Comprehensive resource for GPS, Galileo, GLONASS, BeiDou
  \item
    Topics: Signal structures, receiver design, propagation effects
  \item
    Example: \url{https://gssc.esa.int/navipedia/index.php?title=GALILEO_Signal_Plan}
  \end{itemize}
\item
  \textbf{GPS.gov} - \url{https://www.gps.gov/}

  \begin{itemize}
  \tightlist
  \item
    Official U.S. government GPS information
  \item
    Topics: GPS accuracy, signal specifications, applications
  \end{itemize}
\end{enumerate}

\subsubsection{Spectrum Management}\label{spectrum-management}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{46}
\tightlist
\item
  \textbf{FCC Frequency Allocation Chart} - \url{https://www.ntia.doc.gov/files/ntia/publications/january_2016_spectrum_wall_chart.pdf}

  \begin{itemize}
  \tightlist
  \item
    U.S. spectrum allocations (0-300 GHz)
  \item
    Useful for: Understanding band usage
  \end{itemize}
\item
  \textbf{ITU Radio Regulations} - \url{https://www.itu.int/pub/R-REG-RR}

  \begin{itemize}
  \tightlist
  \item
    International frequency allocations
  \item
    Useful for: Global spectrum planning
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Software Tools}{Software Tools}}\label{software-tools}

\subsubsection{Open Source SDR}\label{open-source-sdr}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{48}
\tightlist
\item
  \textbf{GNURadio} - \url{https://www.gnuradio.org/}

  \begin{itemize}
  \tightlist
  \item
    Open-source SDR toolkit
  \item
    Topics: Signal processing blocks, real-time DSP
  \end{itemize}
\item
  \textbf{SDR++} - \url{https://www.sdrpp.org/}

  \begin{itemize}
  \tightlist
  \item
    Cross-platform SDR receiver software
  \item
    Useful for: Visualizing real-world signals
  \end{itemize}
\item
  \textbf{GQRX} - \url{https://gqrx.dk/}

  \begin{itemize}
  \tightlist
  \item
    Software-defined radio receiver
  \item
    Useful for: Signal monitoring and analysis
  \end{itemize}
\end{enumerate}

\subsubsection{Simulation Tools}\label{simulation-tools}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{51}
\tightlist
\item
  \textbf{Octave/MATLAB Communications Toolbox}

  \begin{itemize}
  \tightlist
  \item
    Commercial and open-source options
  \item
    Topics: Modulation, channel coding, link analysis
  \end{itemize}
\item
  \textbf{Python SciPy/NumPy}

  \begin{itemize}
  \tightlist
  \item
    Free scientific computing libraries
  \item
    Topics: DSP algorithm prototyping
  \end{itemize}
\item
  \textbf{LTspice} - \url{https://www.analog.com/en/design-center/design-tools-and-calculators/ltspice-simulator.html}

  \begin{itemize}
  \tightlist
  \item
    Free circuit simulator
  \item
    Useful for: Analog front-end design
  \end{itemize}
\end{enumerate}

\subsubsection{Antenna Design}\label{antenna-design}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{54}
\tightlist
\item
  \textbf{NEC2} - \url{http://www.nec2.org/}

  \begin{itemize}
  \tightlist
  \item
    Numerical Electromagnetics Code
  \item
    Useful for: Antenna modeling
  \end{itemize}
\item
  \textbf{4nec2} - \url{https://www.qsl.net/4nec2/}

  \begin{itemize}
  \tightlist
  \item
    NEC2 GUI for Windows
  \item
    Useful for: Antenna optimization
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Organizations and Institutions}{Organizations and Institutions}}\label{organizations-and-institutions}

\subsubsection{Professional Societies}\label{professional-societies}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{56}
\tightlist
\item
  \textbf{IEEE Communications Society} - \url{https://www.comsoc.org/}

  \begin{itemize}
  \tightlist
  \item
    Publications: IEEE Trans. on Communications, IEEE Comm. Mag.
  \item
    Conferences: ICC, Globecom
  \end{itemize}
\item
  \textbf{IEEE Signal Processing Society} - \url{https://signalprocessingsociety.org/}

  \begin{itemize}
  \tightlist
  \item
    Publications: IEEE Trans. on Signal Processing
  \item
    Conferences: ICASSP
  \end{itemize}
\item
  \textbf{IEEE Vehicular Technology Society} - \url{https://vtsociety.org/}

  \begin{itemize}
  \tightlist
  \item
    Publications: IEEE Trans. on Vehicular Technology
  \item
    Focus: Mobile and wireless systems
  \end{itemize}
\end{enumerate}

\subsubsection{Research Institutions}\label{research-institutions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{59}
\tightlist
\item
  \textbf{MIT CSAIL (Computer Science and Artificial Intelligence
  Laboratory)} - \url{https://www.csail.mit.edu/}

  \begin{itemize}
  \tightlist
  \item
    Research: Wireless systems, SDR
  \end{itemize}
\item
  \textbf{Bell Labs (Nokia)} - \url{https://www.bell-labs.com/}

  \begin{itemize}
  \tightlist
  \item
    Historic communications research lab
  \item
    Origin of information theory, cellular systems
  \end{itemize}
\item
  \textbf{ESA (European Space Agency)} - \url{https://www.esa.int/}

  \begin{itemize}
  \tightlist
  \item
    Satellite communications and GNSS research
  \item
    Resources: Navipedia, Galileo documentation
  \end{itemize}
\item
  \textbf{NASA JPL (Jet Propulsion Laboratory)} - \url{https://www.jpl.nasa.gov/}

  \begin{itemize}
  \tightlist
  \item
    Deep space communications
  \item
    Research: Advanced coding, large antenna arrays
  \end{itemize}
\end{enumerate}

\subsubsection{Standards Organizations}\label{standards-organizations}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{63}
\tightlist
\item
  \textbf{3GPP (3rd Generation Partnership Project)} - \url{https://www.3gpp.org/}

  \begin{itemize}
  \tightlist
  \item
    LTE, 5G NR specifications
  \end{itemize}
\item
  \textbf{ETSI (European Telecommunications Standards Institute)} - \url{https://www.etsi.org/}

  \begin{itemize}
  \tightlist
  \item
    European telecom standards (DVB, TETRA)
  \end{itemize}
\item
  \textbf{ITU (International Telecommunication Union)} - \url{https://www.itu.int/}

  \begin{itemize}
  \tightlist
  \item
    UN agency for ICT standards
  \item
    Radio Regulations, spectrum management
  \end{itemize}
\end{enumerate}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Appendices Available in Wiki}{Appendices Available in Wiki}}\label{appendices-available-in-wiki}

The following technical documents are included as full appendices in
this wiki:

\begin{itemize}
\tightlist
\item
  \textbf{HRP Framework Paper} - Hyper-Rotational Physics Framework
  (Jones, 2025)
\item
  \textbf{AID Protocol v3.1} - AID Protocol Technical Specification
  v3.1
\item
  \textbf{Biophysical Coupling Mechanism} - Quantum Coherence
  Perturbation Mechanism
\item
  \textbf{Turing Cage} - Turing Cage Analysis
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{\texorpdfstring{Related Wiki Pages}{Related Wiki Pages}}\label{related-wiki-pages}

\begin{itemize}
\tightlist
\item
  \textbf{Home} - Wiki main page
\item
  \textbf{Formula Reference Card} - Quick formula lookup
\item
  \textbf{Glossary} - Terminology reference
\item
  \textbf{Real-World System Examples} - Practical system designs
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\emph{This bibliography is maintained as part of the Chimera DSP Wiki.
Contributions and corrections welcome via GitHub.}
