\newpage
\thispagestyle{empty}

\vspace*{2cm}

\begin{center}
{\Large\bfseries Foreword}
\end{center}

\vspace{1.5cm}

There will come a time in your life when you will ask yourself a series of questions.

\textit{Am I happy with who I am?}

\textit{Am I happy with the people around me?}

\textit{Am I happy with what I'm doing?}

\textit{Am I happy with the way my life is going?}

\textit{Do I have a life or am I just living?}

Do not let these questions strain or trouble you---just point yourself in the direction of your dreams, find your strength in the sound, and make your transition.

\vspace{1cm}

Do not spend too much time thinking and not enough doing.

\textit{Did I try the hardest at any of my dreams?}

\textit{Did I purposely let others discourage me when I knew I could?}

\textit{Will I die never knowing what I could have been or could have done?}

Do not let these doubts restrain or trouble you---just point yourself in the direction of your dreams. Find your strength in the sound and make your transition.

\vspace{1cm}

There will be people who say you can't---you will.

There will be people who say you don't mix this with that, and you will say ``watch me.''

There will be people who will say play it safe, that's too risky---you will take that chance and have no fear.

You won't let these questions restrain or trouble you.

You will point yourself in the direction of your dreams.

You will find the strength in the sound and make your transition.

\vspace{1cm}

For those who know it's time to leave the house and go back to the field:

Find your strength in the sound and make your transition.

\vspace{1.5cm}

\begin{flushright}
\textit{--- Underground Resistance, ``Transition''}
\end{flushright}

\vspace*{\fill}

\newpage
