% ==============================================================================
% CHAPTER 63: Glossary
% ==============================================================================

\chapter{Glossary}
\label{ch:glossary}

\begin{nontechnical}
    This glossary provides a comprehensive quick-reference guide to the terminology, acronyms, and key concepts used throughout this book. It is designed for practitioners, students, and researchers in digital communications and signal processing. Entries are organised for rapid lookup and are cross-referenced to the main chapters where each topic is discussed in detail, creating a conceptual map of the entire work.
\end{nontechnical}

\section{Acronyms and Abbreviations}

\begin{multicols}{3}
\begin{description}
    \item[ACK] Acknowledgment
    \item[ADC] Analogue-to-Digital Converter
    \item[AESA] Active Electronically Scanned Array
    \item[AFC] Automatic Frequency Control
    \item[AGC] Automatic Gain Control
    \item[AID] Auditory Intermodulation Distortion
    \item[AM] Amplitude Modulation
    \item[AMC] Adaptive Modulation and Coding
    \item[AR] Axial Ratio
    \item[ARQ] Automatic Repeat Request
    \item[ASK] Amplitude-Shift Keying
    \item[AWGN] Additive White Gaussian Noise
    \item[BCH] Bose-Chaudhuri-Hocquenghem (code)
    \item[BER] Bit Error Rate
    \item[BPSK] Binary Phase-Shift Keying
    \item[CDF] Cumulative Distribution Function
    \item[CFO] Carrier Frequency Offset
    \item[CQI] Channel Quality Indicator
    \item[CRC] Cyclic Redundancy Check
    \item[DAC] Digital-to-Analogue Converter
    \item[dB] Decibel
    \item[dBi] Decibels relative to Isotropic
    \item[dBm] Decibels relative to 1 milliwatt
    \item[DFE] Decision Feedback Equaliser
    \item[DSSS] Direct-Sequence Spread Spectrum
    \item[DVB-S2] Digital Video Broadcasting - Satellite, 2nd Gen.
    \item[EIRP] Effective Isotropic Radiated Power
    \item[FDD] Frequency-Division Duplex
    \item[FEC] Forward Error Correction
    \item[FER] Frame Error Rate
    \item[FFT] Fast Fourier Transform
    \item[FHSS] Frequency-Hopping Spread Spectrum
    \item[FM] Frequency Modulation
    \item[FSK] Frequency-Shift Keying
    \item[FSPL] Free-Space Path Loss
    \item[GFSK] Gaussian Frequency-Shift Keying
    \item[GNSS] Global Navigation Satellite System
    \item[GPS] Global Positioning System
    \item[HARQ] Hybrid Automatic Repeat Request
    \item[HF] High Frequency
    \item[HRP] Hyper-Rotational Physics
    \item[ICNIRP] Intl. Commission on Non-Ionizing Radiation Protection
    \item[IF] Intermediate Frequency
    \item[IFFT] Inverse Fast Fourier Transform
    \item[IMD] Intermodulation Distortion
    \item[IQ] In-phase and Quadrature
    \item[ISI] Inter-Symbol Interference
    \item[LDPC] Low-Density Parity-Check (code)
    \item[LHCP] Left-Hand Circular Polarisation
    \item[LNA] Low-Noise Amplifier
    \item[LO] Local Oscillator
    \item[LOS] Line-of-Sight
    \item[LPI] Low Probability of Intercept
    \item[LPD] Low Probability of Detection
    \item[LTE] Long-Term Evolution (4G)
    \item[MCS] Modulation and Coding Scheme
    \item[MIMO] Multiple-Input, Multiple-Output
    \item[ML] Maximum Likelihood
    \item[MMSE] Minimum Mean Square Error
    \item[mmWave] Millimetre-Wave
    \item[MSK] Minimum-Shift Keying
    \item[NLOS] Non-Line-of-Sight
    \item[OFDM] Orthogonal Frequency-Division Multiplexing
    \item[OOK] On-Off Keying
    \item[Orch-OR] Orchestrated Objective Reduction
    \item[PA] Power Amplifier
    \item[PAM] Pulse Amplitude Modulation
    \item[PAPR] Peak-to-Average Power Ratio
    \item[PDF] Probability Density Function
    \item[PLL] Phase-Locked Loop
    \item[PSK] Phase-Shift Keying
    \item[QAM] Quadrature Amplitude Modulation
    \item[QCL] Quantum Cascade Laser
    \item[QPSK] Quadrature Phase-Shift Keying
    \item[RF] Radio Frequency
    \item[RHCP] Right-Hand Circular Polarisation
    \item[RS] Reed-Solomon (code)
    \item[RX] Receiver
    \item[SDR] Software-Defined Radio
    \item[SER] Symbol Error Rate
    \item[SINR] Signal-to-Interference-plus-Noise Ratio
    \item[SNR] Signal-to-Noise Ratio
    \item[TDD] Time-Division Duplex
    \item[TEC] Total Electron Content
    \item[THz] Terahertz
    \item[TX] Transmitter
    \item[UHF] Ultra High Frequency
    \item[VHF] Very High Frequency
    \item[WiFi] Wireless Fidelity (IEEE 802.11)
    \item[ZF] Zero-Forcing (equaliser)
\end{description}
\end{multicols}

\begin{importantbox}
\section*{Further Reading: Key Concepts}
This glossary serves as an index to the core concepts of the book. For detailed explanations, derivations, and context for the terms listed above, please refer to the following key chapters:
\begin{description}
    \item[\Cref{ch:baseband-passband} (Baseband vs. Passband)] Defines the fundamental domains of signal processing.
    \item[\Cref{ch:iq} (IQ Representation) and \Cref{ch:constellations} (Constellation Diagrams)] Explain the mathematical and visual language of digital modulation.
    \item[\Cref{ch:snr} (\(SNR\)) and \Cref{ch:energy-ratios} (\(E_b/N_0\))] Detail the essential metrics for signal quality.
    \item[\Cref{ch:ber} (Bit Error Rate)] Quantifies the ultimate performance of a digital link.
    \item[\Cref{ch:shannon} (Shannon's Channel Capacity)] Describes the theoretical limits of communication.
    \item[\Cref{ch:fec} (Forward Error Correction)] Provides an overview of the coding techniques used to achieve reliable communication.
    \item[\Cref{ch:multipath} (Multipath Propagation)] Explains the physics of the wireless channel and the origin of fading.
    \item[\Cref{ch:ofdm} (OFDM) and \Cref{ch:mimo} (MIMO)] Detail the two key technologies that enable modern high-speed wireless systems.
    \item[\Cref{ch:linkbudget} (Complete Link Budget Analysis)] Provides the framework that integrates all these concepts into a practical engineering tool.
    \item[\Cref{ch:orch-or} (Orch-OR) and \Cref{ch:hrp} (HRP Framework)] Detail the speculative but foundational theories of quantum consciousness explored in the latter part of this book.
    \item[\Cref{ch:formula-reference-card} (Formula Reference Card)] For a consolidated list of the most critical equations.
\end{description}
\end{importantbox}