\newpage
\thispagestyle{empty}

\vspace*{3cm}

\begin{center}
{\Large\textsc{Foreword}}
\end{center}

\vspace{2cm}

\begin{center}

There will come a time in your life when you will ask yourself a series of questions.

\vspace{1.2cm}
\textit{Am I happy with who I am?}\\
\textit{Am I happy with the people around me?}\\
\textit{Am I happy with what I'm doing?}\\
\textit{Am I happy with the way my life is going?}\\
\textit{Do I have a life or am I just living?}
\vspace{1.2cm}

\textbf{Do not let these questions strain or trouble you---just point yourself in the direction of your dreams, find your strength in the sound, and \emph{make your transition}}.

\vspace{2cm}

Do not spend too much time thinking and not enough doing.

\vspace{1.2cm}
\textit{Did I try the hardest at any of my dreams?}\\
\textit{Did I purposely let others discourage me when I knew I could?}\\
\textit{Will I die never knowing what I could have been or could have done?}
\vspace{1.2cm}

\textbf{Do not let these doubts restrain or trouble you---just point yourself in the direction of your dreams. Find your strength in the sound and \emph{make your transition}}.

\vspace{2.5cm}

There will be people who say you can't---\textbf{you will.}\\[0.8cm]
There will be people who say you don't mix this with that,\\
and you will say ``\textbf{watch me}.''\\[0.8cm]
There will be people who will say play it safe, that's too risky---\\
\textbf{you will take that chance and have no fear.}\\[1.5cm]

\textbf{You won't let these questions restrain or trouble you.}\\
\textbf{You will point yourself in the direction of your dreams.}\\
\textbf{You will find the strength in the sound and \emph{make your transition}.}

\vspace{2.5cm}

For those who know it's time to leave the house and go back to the field:\\[1cm]
{\Large\textbf{Find your strength in the sound and \emph{make your transition}.}}

\end{center}

\vspace{2cm}

\begin{flushright}
\textit{--- Underground Resistance, ``Transition''}
\end{flushright}

\vspace*{\fill}

\newpage
