\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

\begin{nontechnical}
This book is intended for a wide audience, from students of signal processing to seasoned engineers and researchers. While the foundational chapters provide a solid grounding in established theory, the later sections venture into more speculative and paradigm-challenging territory. These advanced topics are presented not as established fact, but as an invitation to explore the boundaries of our current understanding. They are the result of extensive research and experimentation, and are intended to spark curiosity and debate. We encourage readers to approach this material with an open and critical mind, to question assumptions, and to join us in the exciting quest for new knowledge.
\end{nontechnical}

\vspace{1cm}

The journey you are about to embark on is one of discovery. It begins with the well-trodden paths of digital signal processing (DSP) and wireless communications, but it does not end there. The Chimera project was born from a desire to not only understand the world of signals, but to question its very foundations.

In the following chapters, you will find a comprehensive exploration of the principles that govern our modern connected world. But you will also find something more: a challenge. A challenge to the conventional models, a questioning of long-held assumptions, and a presentation of new ideas that may, at first, seem outlandish.

This is by design. True innovation does not come from accepting the status quo, but from daring to ask "what if?". The advanced concepts presented here, particularly those related to biophysical communication and hyper-rotational physics, are the product of that question. They are presented here not as dogma, but as a starting point for a new conversation.

Whether you are a student seeking to learn, a hobbyist looking to experiment, or a researcher searching for the next frontier, we hope this work provides you with both the tools to build and the inspiration to dream.

Welcome to Chimera.
